Probabilistic inference in dynamical systems

### 2024-04-07 nbs/concepts/inferenceindynamicalsystems/index.qmd
feat(nbs): draft trajectories and phase portraits Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

```{python}
# | label: transcription-splicing-degradation-equations
from sympy import (
    diff,
    Eq,
    Function,
    init_printing,
    solve,
    symbols,
)
from IPython.display import display

@@

```{python}
du_dt_dimless = Eq(
    beta * diff(u_star * alpha / beta, t_star),
    du_dt.rhs.subs(
        {
            u: u_star * alpha / beta,
            s: s_star * alpha / beta,
            t: t_star / beta,
            gamma: gamma_star * beta,
        }
    ),
).simplify()
du_dt_dimless
```

@@

```{python}
ds_dt_dimless = Eq(
    beta * diff(s_star * alpha / beta, t_star),
    ds_dt.rhs.subs(
        {
            u: u_star * alpha / beta,
            s: s_star * alpha / beta,
            t: t_star / beta,
            gamma: gamma_star * beta,
        }
    ),
).simplify()

@@

>>>>

```{python}
#| label: transcription-splicing-degradation-equations
#| code-fold: true
from sympy import (
    diff,
    Eq,
    Function,
    init_printing,
    solve,
    symbols,
)
from IPython.display import display

@@

```{python}
#| code-fold: true
du_dt_dimless = Eq(
    beta * diff(u_star * alpha / beta, t_star),
    du_dt.rhs.subs(
        {
            u: u_star * alpha / beta,
            s: s_star * alpha / beta,
            t: t_star / beta,
            gamma: gamma_star * beta,
        }
    ),
).simplify()
du_dt_dimless
```

@@

```{python}
# | code-fold: true
ds_dt_dimless = Eq(
    beta * diff(s_star * alpha / beta, t_star),
    ds_dt.rhs.subs(
        {
            u: u_star * alpha / beta,
            s: s_star * alpha / beta,
            t: t_star / beta,
            gamma: gamma_star * beta,
        }
    ),
).simplify()

@@

As a simple example, we will simulate the system for a single set of initial
conditions and parameters.

```{python}
# | label: transcription-splicing-degradation-simulation
from pyrovelocity.models import solve_transcription_splicing_model
from jax import numpy as jnp

ts0 = jnp.linspace(0, 4, 40)
ts1 = jnp.linspace(4 + (10 - 4) / 20, 10, 20)
ts = jnp.concatenate([ts0, ts1])
initial_state = jnp.array([0.5, 0.5])
params = (0.5,)

solution = solve_transcription_splicing_model(ts, initial_state, params)
```

We can visualize the results of the simulation by plotting the pre-mRNA and mRNA
concentrations over time as shown in @fig-transcription-splicing-degradation-plot-simulation-trajectory-1,

```{python}
# | label: fig-transcription-splicing-degradation-plot-simulation-trajectory-1
# | code-fold: true
from pyrovelocity.plots import plot_deterministic_simulation_trajectories

plot_deterministic_simulation_trajectories(
    solution,
    title_prefix="TSD Model",
)
```

and the phase portrait in @fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1.

```{python}
# | label: fig-transcription-splicing-degradation-plot-simulation-phase-portrait-1
# | code-fold: true
from pyrovelocity.plots import plot_deterministic_simulation_phase_portrait

plot_deterministic_simulation_phase_portrait(
    solution,
    title_prefix="TSD Model",
)
```
